Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-18 paper/paper.tex
Refs
Introduction with my ref

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity.  For recursion, this is done by automatically inferring
which argument in the recursive call gets smaller with respects to the original
input argument.  For productivity, the algorithm checks that the corecursive
call appears directly under a constructor to make sure that this function
always produces at least one element after each recursive step.
This implies that some functions, though well-defined, cannot be accepted by
the proof assistant. Examples of this include common sorting algorithms, such
as quicksort and mergesort.
%One example is Quicksort, written in OCaml
%below:
\begin{minted}{OCaml}
let rec qsort xs = match divide xs with | None -> []
 | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input ``dives'' deeper in the input returning
two sublists with the head as a pivot.

The main approach for implementing nonstructural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are coupled with
\emph{termination proofs}.  Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
one approach is using the \coq{Fix} combinator, in which recursion is done on an
\emph{accesibility} predicate on the input, which is a proof that there are no
infinitely decreasing chains.}. A benefit of this approach is that, extracting
verified nonstructural recursive functions to OCaml will erase the termination
proofs, and so produce code that will be closer to what a programmer may have
written directly in OCaml. However, reasoning about program equivalences
requires dealing with such termination proofs, and it is common practice to use
custom reduction lemmas that will be used extensively when proving properties
about nonstructural recursive functions.

\emph{Structured recursion schemes} have been successfully employed in
functional programming to structure recursive programs, for example, quicksort
and mergesort are both instances of a divide-and-conquer algorithm which can be
formalised using hylomorphisms~\cite{MeijerFP91, HuIT96}. In fact,  Hinze et
al.~\cite{HinzeWG15} showed that \emph{every recursive program can be formalised
as a hylomorphism}.  Furthermore, hylomorphism laws can capture a number of
useful equivalences, ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to semi-automatic
parallelisations~\cite{Gibbons96:Third, farmsCastro}.
% Recursion schemes have successfully been applied to functional programming to
% structure recursive programs, however, because of the presence of
% non-termination in languages like (e.g.) Haskell, they cannot guarantee
% recursive definitions well-defined. % This is due to the possibility of applying a
% %
% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
% % use of data functors.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow the extraction of idiomatic OCaml code.  By this we refer to code that
(1) does not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code.  This latter is problematic in that, for higher-order programs, simple interoperations can lead to incorrect behaviour or
even segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates
the fast-and-loose principle~\cite{DanielssonHJG06}. % The absence
% of \ocaml{Obj.magic} provides even stronger
% guarantees to the extracted code. In particular, the extracted code from Coq is
% well-defined allowing the programmer to use the so-called \emph{fast-and-loose
% reasoning}~\cite{DanielssonHJG06} on the extracted code. % For example, the
% % programmer would need to manually
% %
% check that the inputs satisfy the necessary assumptions which would be
% % impossible to enforce in the presence of unsafe casts. Overall, the absence
% % of \ocaml{Obj.magic} makes it easer to reason about, and interoperate with
% % extracted code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2) allows the
extraction of idiomatic OCaml code. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code. As an example our mechanisation of \ocaml{qsort} as a
hylomorphism will produce the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}

Finally, in this paper we provide a framework for \emph{generic programming} with recursion
schemes in Coq, with a proof of their algebraic laws and program equivalences for clean code extraction:
\begin{itemize}
  \item In Section~\ref{sec:containers} we formalise the type of container functors suitably modified for program extraction
  \item In Section~\ref{sec:recursion-schemes} we formalise hylomorhisms and prove their universal property
  \item In Section~\ref{sec:examples} we provide examples of divide and conquer, dynamic programming, and mutual recursion algorithms and we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
\end{itemize}

>>>>

\section{Introduction}
\label{sec:intro}
Recursive definitions cannot be proven well-defined automatically due to the
halting problem. Modern proof assistants like Coq or Agda provide a sound, but
incomplete algorithm which syntactically checks for termination or
productivity.  For recursion, this is done by automatically inferring
which argument in the recursive call gets smaller with respects to the original
input argument.  For productivity, the algorithm checks that the corecursive
call appears directly under a constructor to make sure that this function
always produces at least one element after each recursive step.
This implies that some functions, though well-defined, cannot be accepted by
the proof assistant. Examples of this include common sorting algorithms, such
as quicksort:
%One example is Quicksort, written in OCaml
%below:
\begin{minted}{OCaml}
let rec qsort xs = match divide xs with | None -> []
 | Some (pivot, (smaller, larger)) -> qsort smaller @ (pivot::qsort larger)
\end{minted}
While \ocaml{qsort} is a well-defined mathematical function it cannot be
accepted by a proof assistant.  The reason is that the \ocaml{divide}
function \emph{destructuring} the input dives deeper in the input returning
two sublists with the head as a pivot.

The main approach for implementing non-structural recursion in Coq is to use
\emph{well-founded recursion}, where recursive definitions are coupled with
\emph{termination proofs}.  Using well-founded recursion, the recursive calls
will happen on \emph{structurally smaller termination proofs}\footnote{In Coq,
one approach is using the \coq{Fix} combinator, in which recursion is done on an
\emph{accesibility} predicate on the input, which is a proof that there are no
infinitely decreasing chains.}. A benefit of this approach is that, extracting
verified nonstructural recursive functions to OCaml will erase the termination
proofs, and so produce code that will be closer to what a programmer may have
written directly in OCaml. However, reasoning about program equivalences
requires dealing with such termination proofs, and it is common practice to use
custom reduction lemmas that will be used extensively when proving properties
about nonstructural recursive functions.

\emph{Structured recursion schemes} have been successfully employed in
functional programming to structure recursive programs, for example, quicksort
and mergesort are both instances of a divide-and-conquer algorithm which in
terms of recursion schemes can be can be formalised as an
hylomorphism~\cite{MeijerFP91, HuIT96}. Furthermore, it has been shown by Hinze
et al.~\cite{HinzeWG15} that hylomorphisms provide the basic building block of
every recursion scheme. In particular, any complex recursion scheme can be
transmogrified down into an hylomorphisms by means of an adjunction.
Furthermore, hylomorphism laws can capture a number of useful equivalences,
ranging from common optimisations such as \emph{short-cut
fusion}~\cite{TakanoM95}, to semi-automatic
parallelisations~\cite{Gibbons96:Third, farmsCastro}.
% Recursion schemes have successfully been applied to functional programming to
% structure recursive programs, however, because of the presence of
% non-termination in languages like (e.g.) Haskell, they cannot guarantee
% recursive definitions well-defined. % This is due to the possibility of applying a
% %
% recursion scheme to a non-terminating (co)algebra, but also to the unrestricted
% % use of data functors.

@@

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow for extraction of idiomatic OCaml code. This is problematic since code
extration has proven useful in a variety of scenarios~\cite{OnoHTNH11,
Larchey-Wendling23, MiculanP12, Sakaguchi20}.  However, most uses of extraction
(1) do not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code. This latter
is also problematic in that, for higher-order programs, simple interoperations
can lead to incorrect behaviour or even segfaults~\cite{forster:hal-04329663}
and, moreover, it invalidates the fast-and-loose
principle~\cite{DanielssonHJG06}.

This work presents the first formalisation in Coq of \emph{hylomorphisms} that (1) is \emph{axiom-free} and (2) allows the
extraction of idiomatic OCaml code. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code (see extracted \ocaml{qsort} program in Section~\ref{sec:examples}). % As an example, our mechanisation of \ocaml{qsort} as a
% hylomorphism will produce the following code:
% \begin{coqcode}
% let rec qsort = function | [] -> []
%   | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
%               let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in
%               app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
% \end{coqcode}

To summarise, our contributions are as follows. In this paper we provide a
framework for \emph{generic programming} with recursion schemes in Coq, with a
proof of their algebraic laws and program equivalences for clean code
extraction:
\begin{itemize}
  \item In Section~\ref{sec:containers} we formalise the type of container functors ensuring the presence of least and greatest fixed-points for functors and suitably adapted for program extraction
  \item In Section~\ref{sec:recursion-schemes} we formalise folds, unfolds and hylomorhisms and their universal properties
  \item In Section~\ref{sec:examples} we use the framework to formalise examples of divide-and-conquer, dynamic programming, and mutual recursion algorithms. Furthermore we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
\end{itemize}
